Loan Data from Prosper
Last updated 03/11/2014

#Introduction 

Prosper is America’s first marketplace lending platform, with over $10 billion in funded loans.Prosper allows people to invest in each other in a way that is financially and socially rewarding.This data set contains 113,937 loans with 81 variables on each loan, including, borrower rate , loan amount, loan status, borrower income, borrower employment status,ProsperScore, Loan Current Days Delinquent and other details.

========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}


library(knitr)
library(ggplot2)

```

```{r echo=FALSE,warning=FALSE,message=FALSE, Load_the_Data}
# Load the Data
loan_data=read.csv("prosperLoanData.csv")

```



# Univariate Plots Section


```{r echo=FALSE,warning=FALSE,message=FALSE, Univariate_Plots1}
#Employment Status

ggplot(aes(x = EmploymentStatus), data = loan_data) + 
  geom_bar(fill = "#C4E538") + 
  theme(axis.text.x = element_text(angle = 90)) + 
  xlab("Employment Status")+
  ylab("Number of Employment status")

summary(loan_data$EmploymentStatus)
```

we notice that the employed category has the best opportunity of loan unlike unemployed

```{r echo=FALSE,warning=FALSE,message=FALSE, Univariate_Plots2}
ggplot(aes(x = ProsperScore), data = loan_data) + 
  geom_bar(fill = '#C4E538') +
  scale_x_continuous(breaks = 1:11) +
  xlab("Prosper Score")

summary(loan_data$ProsperScore)
```

This is plot shows that it is normally distributed and the mean of prosper score =5.95

```{r echo=FALSE,warning=FALSE,message=FALSE, Univariate_Plots3}
#LoanStatus
ggplot(aes(x=LoanStatus),data = loan_data)+
  geom_bar(fill="#C4E538")+
  theme(axis.text.x = element_text(angle = 90))+
  xlab("Loan Status")+
  ylab("Number of loan status")

summary(loan_data$LoanStatus)
```

we notice that the majority of the loan status are either current or completed 
in the other side there is no Cancelled.

                
```{r echo=FALSE,warning=FALSE,message=FALSE, Univariate_Plots4}

ggplot(aes(x=IncomeRange),data = loan_data)+
  geom_bar(fill="#C4E538")+
  theme(axis.text.x = element_text(angle = 90))
  xlab("Income Range")

  summary(loan_data$IncomeRange)
```
The income ranges show that the vast majority of borrowers make between 25,000 and 75,000 per year
```{r echo=FALSE,warning=FALSE,message=FALSE, Univariate_Plots5}
loan_data$Term.f <- as.factor(loan_data$Term)
ggplot(aes(x=Term.f),data = loan_data)+
  geom_bar(fill="#C4E538")+
  xlab("Loan in months")

summary(loan_data$Term)
```

the mean of the period of the loan is 36 months
  
```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(x=LoanCurrentDaysDelinquent),data = loan_data)+
  geom_histogram(fill="#C4E538")+
  xlab("Loan Current Days Delinquent")+
  scale_x_log10()
summary(loan_data$LoanCurrentDaysDelinquent)
```

we notice that the Days Delinquentof the most of borrowers are 152.8 days

```{r echo=FALSE,warning=FALSE,message=FALSE}
loan_data$LoanOriginationDate <- as.Date(loan_data$LoanOriginationDate,format = "%Y-%m-%d")
loan_data$LoanOriginationDate.year=format(loan_data$LoanOriginationDate, "%Y")
ggplot(aes(x = LoanOriginationDate.year), data = loan_data) +
  geom_bar(fill="#C4E538") + 
  ggtitle('Number of Loans by Year')
```

the biggest amount of loans was in 2013



```{r echo=FALSE,warning=FALSE,message=FALSE}

ggplot(aes(x = EmploymentStatusDuration/12), data = loan_data) + 
  geom_histogram(binwidth=1, fill = "#C4E538") +
  ggtitle("Employment Duration of the borrowers") +
  xlab("Employment duration (years)") +
  theme(text = element_text(size=12))
```
 
 By increasing the durations of the work, the loan will be decreased
 
```{r echo=FALSE,warning=FALSE,message=FALSE}
#Is Borrower Homeowner 

ggplot(aes(x = IsBorrowerHomeowner), data = loan_data) + 
  geom_bar(fill = "#C4E538") + 
  xlab("Is Borrower Homeowner ")

summary(loan_data$IsBorrowerHomeowner)
```

The number of people who own  houses almost equal The number of people who don't own houses

```{r echo=FALSE,warning=FALSE,message=FALSE}
#DebtToIncomeRatio

ggplot(aes(x = DebtToIncomeRatio), data = loan_data) + 
  geom_histogram(fill = "#C4E538")+
  scale_x_continuous(lim=c(0,quantile(loan_data$DebtToIncomeRatio,0.99,na.rm=TRUE)))+
  xlab("DebtToIncomeRatio")

summary(loan_data$DebtToIncomeRatio)
```

```{r echo=FALSE,warning=FALSE,message=FALSE}
#How many borrowers had defaulted? Particularly interesting was between 2007 - 2009 i.e. the mortgage crisis era
loan_data$LoanOriginationDate.year.num <- as.numeric(loan_data$LoanOriginationDate.year)
ggplot(aes(x = LoanStatus),
       data = subset(loan_data,loan_data$LoanOriginationDate.year.num >= 2007 & loan_data$LoanOriginationDate.year.num <= 2009)) + 
  geom_bar(fill = "#C4E538")+
  xlab("Loan Status between 2007 - 2009 ")

```

the number of borrowers who had defaulted between 2007 - 2009 was 5928


# Univariate Analysis

### What is the structure of your dataset?

This is dataset contains 113937 loans and 81 variables for Prosper Marketplace

### What is/are the main feature(s) of interest in your dataset?
####- Income range
####- employment Status
####- loan Status
####- ProsperScore
####- Term
####- Loan Current Days Delinquent
####- Is Borrower Homeowner
####- Borrower Rate
####- Loan Year

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
####- occuption 
####- Employment Duration
####- DelinquenciesInLast7Years
####- BankCard Utilization


### Did you create any new variables from existing variables in the dataset?
-LoanOriginationDate.year => to define the year of the loan

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

-Outliers handling  for Loan Current Days Delinquent by log-scaling (scale_x_log10()) 

# Bivariate Plots Section



```{r echo=FALSE,warning=FALSE,message=FALSE}
#Does the employment status effect the borrower rate?

ggplot(aes(x = Term.f, y = BorrowerRate), data = loan_data) + 
  geom_boxplot() + 
  xlab("Term") +
  ylab("Borrower Rate")
loan_data$Term.months <- as.numeric(loan_data$Term)
cor.test(loan_data$Term.months,loan_data$BorrowerRate)
```

the period of loan doesn't affect the borrower rate

```{r echo=FALSE,warning=FALSE,message=FALSE}

ggplot(aes(x = EmploymentStatus, y = BorrowerRate), data = loan_data) + 
  geom_boxplot(aes(fill = EmploymentStatus), show.legend = F) + 
  theme(axis.text.x = element_text(angle = 90))+
  xlab("Employment Status") +
  ylab("Borrower Rate")

```

the unemploymed status has the higher borrower rate 

```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(y = BorrowerRate, x = ProsperRating..numeric.), 
       data = subset(loan_data, loan_data$StatedMonthlyIncome < 10000 &
                       !is.na(ProsperRating..numeric.))) +
  geom_point(alpha = 1/20, position = position_jitter( h = 0))

with(subset(loan_data, loan_data$StatedMonthlyIncome < 10000 &
               !is.na(ProsperRating..numeric.)),
     cor.test(ProsperRating..numeric., BorrowerRate,
              method = 'pearson'))
```

we notice that the loans with higher interest rate are much more common in loans with higher risk (smaller scores)

-There is a strong negative relationship between Prosper Rating and interest rate

```{r echo=FALSE,warning=FALSE,message=FALSE}


ggplot(aes(x = IsBorrowerHomeowner, y =BorrowerRate),
       data = subset(loan_data,BorrowerRate<0.39 & !is.na(DebtToIncomeRatio) )) +
  scale_color_brewer(type = 'div') +
  geom_boxplot(color = "#C4E538") +
  ggtitle('BorrowerRate when Borrower is Home Owner')

with(subset(loan_data,BorrowerRate<0.39 & !is.na(DebtToIncomeRatio) ),
     cor.test(ProsperRating..numeric., BorrowerRate,
              method = 'pearson'))
```

if the borrower has a home ,the interest rate will be decreased



```{r echo=FALSE,warning=FALSE,message=FALSE}

ggplot(aes(x = LoanOriginalAmount, y = Investors), data = loan_data) + 
  geom_point(alpha = 1/20) +
  geom_smooth(method = 'lm', color = '#C4E538') +
  ggtitle("Loan Original Amount vs Number of Investors") +
  xlab("Loan Original Amount")


```

if the inverstors increase, the amount of loan will increase

```{r echo=FALSE,warning=FALSE,message=FALSE}

ggplot(aes(x = IncomeRange, y = LoanOriginalAmount), data=loan_data) +
    geom_boxplot() +
  ggtitle("Income Range  vs Loan Amount") +
  xlab("Income Range") +
  theme(text = element_text(size=12), 
        axis.text.x = element_text(angle=90, vjust=1)) +
  ylab("Loan Original Amount")


```

larger incomes  on average  Larger loans.

```{r echo=FALSE,warning=FALSE,message=FALSE}

ggplot(aes(x = LoanOriginationDate.year, y =LoanOriginalAmount),
       data = loan_data) +
  scale_color_brewer(type = 'div') +
  geom_boxplot(color = "#C4E538") +
  ggtitle('loan amount grown over the years')


```

loan amount grown over the years particularlly from 2009 to 2014

# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

- larger incomes  on average  Larger loans.
- loan amount grown over the years particularlly from 2009 to 2014
- if the inverstors increase, the amount of loan will increase

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
- By increasing the durations of the work, the loan will be decreased

### What was the strongest relationship you found?

the unemploymed status has the higher borrower rate 


# Multivariate Analysis

```{r echo=FALSE,warning=FALSE,message=FALSE, Bivariate_Plots}

ggplot(aes(x = Term.f, y = LoanOriginalAmount,fill="EmploymentStatus"), data = loan_data) + 
  geom_boxplot(aes(fill = EmploymentStatus)) + 
  xlab("Term") +
  ylab("Loan Original Amount")

cor.test(loan_data$Term.months,loan_data$LoanOriginalAmount)

```

- When the period is increased the amount of loan is increased
- The correlation coefficie indicate a moderate positive linear relationship between amount of loan and period.

```{r echo=FALSE,warning=FALSE,message=FALSE}
loan_data$DebtToIncomeRatio.bucket <- 
  cut(loan_data$DebtToIncomeRatio,  breaks = c(0,0.14,0.22,0.31,1, 2, 4, 6, 10.2))
loan_data$Term.f <- as.factor(loan_data$Term)

ggplot(aes(x = DebtToIncomeRatio.bucket, y =BorrowerRate, color = Term.f ),
       data =  subset(loan_data, !is.na(DebtToIncomeRatio.bucket))) +
  theme(legend.position = "bottom") +
  geom_point( position = position_jitter( h = 0)) +
  scale_color_grey() +  
  ggtitle('BorrowerRate by Term and Debt to Income Ratio') 
```



### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

--Employed borrowers recorded a higher proportion of completed loans, followed by retired borrowers

--There is a strong positive relationship between term and loan amount (longer the term, the larger the loan)


### Were there any interesting or surprising interactions between features?

- In general there is no surprises

------

# Final Plots and Summary

### Plot One

```{r echo=FALSE,warning=FALSE,message=FALSE, Plot_One}

ggplot(aes(x = ProsperRating..Alpha., fill = LoanStatus), data = loan_data )+ 
    geom_bar(position = "fill") +
  labs(x = "Prosper Rating", y = "% of loans", title = "Loan vs Prosper Rating") +
  scale_fill_discrete(name = "Loan Status")
```

### Description One

when loans rated AA have a significantly higher proportion of completed loans. 
This means you can invest in loans rated AA with relatively low risk.

### Plot Two

```{r echo=FALSE,warning=FALSE,message=FALSE, Plot_Two}

ggplot(aes(x = BorrowerRate, y = EstimatedLoss), data = loan_data) +
  geom_jitter(aes(colour = ProsperRating..Alpha.), position = "identity") +
  geom_smooth() +
  labs(x="Borrower Rate", y="Estimated Loss", title="Borrower Rate by Estimated Loss") + 
  scale_color_discrete(name="Prosper Rating")
```

### Description Two

the borrower rate increasing with esimated loss and higher risk loans. 
While the Borrower Rate  increases for loans considered to be higher risk

### Plot Three

```{r echo=FALSE,warning=FALSE,message=FALSE, Plot_Three}
ggplot(aes(x = IncomeRange, y = LenderYield), data = loan_data) + 
  geom_boxplot(aes(fill = IncomeRange), show.legend = F) +
  labs(x="Income Range", y="Lender Yield") +
  theme(axis.text.x = element_text(angle = 90)) 
```

### Description Three

 income range of $1 - $ 24,999 recorded the highest lender yield
 
------

# Reflection

I selected the loan data to test my understanding of the course because it is bigger than the other datasets. I face some problems like dealing with outlires and ggplot syntax, but with the course materials and google, all problems are solved.
Finally, I am trying to make a predictive models, such as a logistic regression for loan origination amount, Income Range,Loan Current Days Delinquent and Prosper score.